Cluster Difference Imaging Photometric Survey. IV. TOI 1937Ab. Youngest Hot Jupiter, or Just Tidally Spun Up?

### 2021-02-22 cluster_paper/ms.tex
update paper
the introduction is no longer garbage

\defcitealias{bouma_wasp4b_2019}{B19}
\defcitealias{cantatgaudin_gaia_2018}{CG18}
\defcitealias{kounkel_untangling_2019}{KC19}

@@

% Context, aims, methods, results, conclusions.
\begin{abstract}
  Recent analyses of the Gaia data have reported the existence of
  diffuse stellar populations surrounding nearby open clusters. 
  The stars in these diffuse halos could have escaped from the cores of the
  cluster, or they could be dispersed remnants of the initial star formation complex.
  They could also be field stars---the false positives of overly
  aggressive clustering algorithms.
  In this study, we focus on a halo with a length of 300\,pc reported
  around the $\approx$120\,Myr open cluster NGC\,2516.
  %TODO: says who?
  The classical tidal radius of the cluster is $\approx$10\,pc.
  Combining photometry from Gaia, rotation periods from TESS, and
  lithium measurements from Gaia-ESO and GALAH, we find that the halo
  of NGC\,2516 is real and coeval with the core.
  Two in three kinematically selected halo members 25-250\,pc from the
  cluster center show rotation periods consistent with a
  gyrochronological age of 120\,Myr.
  A comparison sample of field stars shows no such trend.
  The rotation period and lithium abundances of the cluster stars are
  also anti-correlated at fixed stellar mass, as has been observed in
  similar stellar populations.
  This work expands the set of confirmed NGC\,2516 members by a factor
  of $\approx$2, and quantifies the degree to which Gaia-based
  analyses can identify dispersed stellar populations against a
  background of field star contaminants.
  We highlight the implications for spectroscopic survey targeting,
  open cluster dispersal, and planet searches around young stars.
\end{abstract}

@@

Star clusters form within hierarchically structured molecular clouds
(CITE).  Over the first $\approx$10\,Myr, ``feedback'' effects
including protostellar outflows, photoionization, radiation pressure,
and supernova shocks disperse the gas out of the cloud (CITE
Krumholz+19).  Since only a small fraction of the cloud mass is
converted into stars ($\approx 1\%$; see Krumholz+19 Figure 10), the
gas dispersal drives the majority ($\approx$90\%) of stars in the
cluster to become gravitationally unbound (e.g., Lada \& Lada 03).

From 10 to 1000\,Myr, the portion of the cluster that survives gas
dispersal suffers an onslaught that typically unbinds the remaining stars.
Internal to the cluster, mass loss from supernovae and AGB winds, and
close stellar encounters (``relaxation'') are important over 10s and
100s of megayears, respectively (CITE, CITE).  External to the
cluster, collisions with GMCs (Spitzer 1958), and perturbations from
the galactic tide in both the radial (CITE) and vertical (CITE)
dimensions further drive stars toward the ``high velocity tails'' of
the stellar energy distribution.

% In the traditional picture of star formation molecular clouds go
% gravitationally unstable, and collapse into little knots.  The knots
% produce groups of stars close to each other.  The resulting ``open star
% clusters'' are, as with any first-order guess, spherical.

% How does the cluster evolve?  The answer depends on the total stellar
% mass.  The smallest 90\% of star clusters disperse during the ``embedded
% phase'' (CITE Lada \& Lada 03, Krumholz 19 ARAA).  This is driven mostly
% by {\it process A} (CITE).  The more massive clusters make it maybe 100
% Myr. Maybe a bit longer.  Their evaporation is thought to be driven by
% collisions with molecular clouds (CITE, Spitzer 1958, maybe Sec 7.2 of
% Ryden), and the galactic tide (CITE), and {\it process A}, and {\it
% process B}.  Of course, the expected evaporation time should depend on
% factors including the mass of the cluster itself, and how many high-mass
% (O and B) stars form, since their winds [and maybe supernovae] clear out
% most of the cloud, and the initial density of the gas cloud to begin
% with.
% 
% Assuming that the cluster achieves virial equilibrium, stellar fly-bys
% then conspire to segregate the stellar mass distribution within the
% cluster, evaporating the lowest mass stars soonest (2T+U=0, U=-GM/r,
% assuming equipartition of the specific ``thermal energy'' per star, then
% kT = 0.5$mv^2$ at lower masses requires higher velocities).

Identifying the stars that have dispersed from their natal clusters
into the galactic field is of interest for a few reasons.  One 
is to understand the predominant conditions under which stars and star
clusters form (e.g., Wright Mamajek 2018 ScoCen), in addition to
the processes that drive cluster dissolution.
Another is to understand the Sun's birth conditions---
did it form in an open
cluster, and if so, what type? Is there any hope at
identifying the stars that formed near the Sun (e.g., Adams 2010)?
Finally, the dispersal of open clusters also provides 
a strong test case for ``chemical tagging'', aka.\
``galactic archaeology'' for stars in the thin and thick disk of
the Galaxy (CITE).  If one cannot chemically tag stars
that are known to be kinematically associated, why would 
it be possible to do so once
they are radially mixed throughout the galactic disk?

Identifying the remnant
halos\footnote{ The low-density outer regions of open clusters have
been referred to in general terms as ``halos'' and ``coronae''; we
adopt the former (CITE, CITE). } of open clusters is also important for a
separate project: that of discovering young transiting planets.  Young
transiting planets are hard to find because young stars are rare
(e.g., CITE Piskunov 2018), and reside in crowded regions of the
sky (e.g., Kharchenko 2013).  If the dispersed halos of
nearby star clusters could be reliably identified in a complete
manner, this could expand the census of nearby young stars by up to
a factor of $10$, based on the expected fraction of stars thought to
be lost during gas dispersal.

The identification of low-density (GIVE A NUMBER DENSITY) stellar
associations, though possible before Gaia (e.g.,
Zuckerman+ CITE, Bergond+01), has reached a breakthrough pace in the present era.
Some low-density moving groups have no discernable
cores (e.g., Psc Eri CITE, mu Tau CITE); others are hierarchically
sub-structured (e.g., USco CITE, Taurus[?]).
The focus of this work is
low-density
halos that are associated with previously known open clusters (e.g.,
Kounkel+19, Kounkel+20, Meingast+21).  In some
cases, these halos correspond to tidal tails, e.g., in the Hyades
(CITE Meingast+Alves 20, Roser+20), and in the Ursa Major moving
group (CITE Gagne+20). 
However, applying different clustering methods on the
Gaia data yields different results (CITE: Hunt \& Reffert 2020).
Using say Gaussian Mixture Modelling, or any analogous method that
requires ``clusters'' to be ellipses in phase-space unsurprisingly
yields open clusters that are roughly elliptical.  Some unsupervised
clustering methods lead to similar results \citep[][hereafter
\citetalias{cantatgaudin_gaia_2018}]{cantatgaudin_gaia_2018}.  Other
unsupervised methods
have yielded dispersed and asymmetric structures
with number densities down to 100 times less than the field, spread
over many hundreds of parsecs
(e.g., \citep[][hereafter
\citetalias{kounkel_untangling_2019}]{kounkel_untangling_2019}; Meingast+21).

% TODO: make the bonafide number below quantitative
We've recently been making TESS light curves of age-dated stars across
the sky, as part of a Cluster Difference Imaging Photometry Survey
(CDIPS, CITE Bouma+19).  Our analysis of Cycle 1 (Sectors 1-13) yielded
light curves of \numsouthernuniqlcs candidate cluster members in the
Southern Ecliptic hemisphere, available on
MAST\footnote{\url{https://archive.stsci.edu/hlsp/cdips}}.  Based on
rotation periods, $\approx$25\% appear to be bonafide cluster members.

As part of a broader project of identifying a large and clean sample of
young stars for a transit search, we focus in this paper on a 
modest question: in just a single rich southern open cluster, is the
cluster halo coeval with the core?  We chose
NGC\,2516 for this analysis since it is young ($\sim$120\,Myr; B3V at
the turnoff) and close
($d\approx400$\,pc) enough to facilitate rotation measurements using
TESS, and some degree of spectroscopic analysis.  
Its halo, spanning $\approx$300\,pc, is also quite large.
We want to know: is
the halo real? To what extent can we use Gaia alone to reliably identify
age-dated needles in the haystack of boring field stars?  And more
generally, what are the implications for the evolution of open clusters
if they do have halos?

Section~\ref{sec:gaia} presents the astrometric data from Gaia, and
clarifies our usage of the terms ``core'' and ``halo''.
Section~\ref{sec:agedate} age-dates the halo of NGC\,2516, based on Gaia
photometry (Section~\ref{subsec:hr}), TESS gyrochronology
(Section~\ref{subsec:tess}), and lithium depletion
(Section~\ref{subsec:lithium}).  In Section~\ref{sec:discussion} we
discuss the implications of this analysis for NGC\,2516 specifically and
stellar spin-down and open cluster evolution generally.
Section~\ref{sec:conclusion} gives our conclusions.

@@

Gaia-ESO selected {\it cluster candidates} to be observed with the
GIRAFFE and UVES spectrographs based on previously reported literature
members and publicly available optical and NIR photometry.  This
target selection approach included many stars in the spatial proximity
of previously known members.  Since the possible existence of the \cn
halo was not known at the time of targetting, very few ``halo'' stars
are in their ultimate sample.  For \cn, \citet{randich_gaiaeso_2018}
ultimately reported stellar parameters (including lithium equivalent
widths and metallicities) for 796 candidate members.  Cross-matching
against our kinematic list of \nkinematic candidate members by position and
imposing a 0$\farcs$5 maximum separation limit yields 492 kinematic
members for which a Gaia-ESO spectrum was acquired and stellar
parameters were reported.  15 of these (all with $(Bp-Rp)_0 > 2.0$)
are spurious matches based on the Gaia color and GES effective
temperature, and we remove them.  This yields 477 stars, of which 436
are in the core, and 41 are in the halo.  This ratio is primarily
caused by the Gaia-ESO targetting selection function.

@@

% TODO: fix the fact that the figure currently only shows Randich+18
Figure~\ref{fig:lithium} shows the results.  At fixed stellar mass
(and age), the rapid rotators tend to show elevated lithium equivalent
widths. This effect is mostly apparent in the K dwarfs.  Similar
trends in the Pleiades were noted nearly three decades ago by
\citet{soderblom_evolution_1993}.  More recent studies have found
comparable results in the Pleiades, the Psc-Eri stream, and M\,35
\citep{bouvier_pleiades_lirot_2018,arancibia_2020,jeffries_m35_li_2020}.
The fact that we see the same trend across the core and halo of \cn
{\it a)} supports the conclusion that the halo is coeval with the
core, and {\it b)} suggests that the lithium-rotation correlation is
not explained by environmental effects such as the density of the cluster
during its embedded phase, but is instead tied to physical processes
acting on all low-mass pre-main sequence stars.

>>>>

\defcitealias{bouma_wasp4b_2019}{B19}
\defcitealias{cantatgaudin_gaia_2018}{CG18}
\defcitealias{kounkel_untangling_2019}{KC19}
\defcitealias{meingast_2021}{M21}

@@

% Context, aims, methods, results, conclusions.
\begin{abstract}
  Recent analyses of the Gaia data have reported the existence of
  diffuse stellar populations surrounding nearby open clusters. 
  The stars in these diffuse halos could be the tidal escapees of the
  cluster cores, or they could be remnants of an initially
  dispersed star formation complex.
  They could also be field stars---the false positives of overly
  aggressive clustering algorithms.
  In this study, we focus on a halo with a length of 300\,pc reported
  around the $\approx$120\,Myr open cluster NGC\,2516.
  %TODO: says who?
  The classical tidal radius of the cluster is $\approx$10\,pc.
  Combining photometry from Gaia, rotation periods from TESS, and
  lithium measurements from Gaia-ESO and GALAH, we find that the halo
  of NGC\,2516 is real and coeval with the core.
  Two in three kinematically selected halo members 25-250\,pc from the
  cluster center show rotation periods consistent with a
  gyrochronological age of 120\,Myr.
  A comparison sample of field stars shows no such trend.
  The rotation period and lithium abundances of the cluster stars are
  also anti-correlated at fixed stellar mass, as has been observed in
  similar stellar populations.
  This work expands the set of confirmed NGC\,2516 members by a factor
  of $\approx$2, and quantifies the degree to which Gaia-based
  analyses can identify dispersed stellar populations against a
  background of field star contaminants.
  We highlight the implications for spectroscopic survey targeting,
  open cluster dispersal, and planet searches around young stars.
\end{abstract}

@@

Star clusters form in hierarchically structured molecular clouds
\citep{shu_star_1987}.  Over the first 10\,Myr, ``feedback'' effects
including protostellar outflows, photoionization, radiation pressure,
and supernova shocks disperse the gas out of the cloud
\citep{krumholz_star_2019}.  Since only a small fraction of the cloud
mass is converted into stars ($\sim 1\%$), gas dispersal enables the
majority ($\sim$90\%) of stars in the cluster to escape the cluster's
gravitational well \citep{lada_embedded_2003}.

From 10 to 1000\,Myr, the cluster remnants that survive gas dispersal
suffer an onslaught that almost always leads to dissolution.
Within the cluster supernovae, AGB winds, and close stellar encounters
lead to additional mass loss \citep{lamers_mass_loss_2010}.  Extrinsic to
the cluster, collisions with giant molecular clouds
\citep{spitzer_disruption_1958}, and perturbations from the galactic
tide in both the radial and vertical dimensions further promote
stellar escape \citep[{\it
e.g.},][]{fukushige_timescale_2000,bergond_gravitational_2001}.

Finding the stars that have dispersed from their natal clusters
into the galactic field is of interest for a few reasons.  One is to
understand the predominant conditions under which stars and star
clusters form \citep[{\it e.g.},][]{wright_kinematics_2018}.  Another is
to understand the Sun's birth environment
\citep{adams_birth_2010}.  Did the Sun form in an open cluster,
and if so, what type? Is there any hope at identifying the stars that
formed near the Sun?  Open clusters and their dispersed
remnants also provide a test case for ``chemical tagging'' \citep[{\it
e.g.},]{freeman_new_2002,hogg_chemical_2016,ness_dopplergangers_2018}.
The converse holds too: one could use chemical tagging to test the
viability of kinematic clustering methods.

A separate project that benefits from the new discoveries of dispersed
stellar populations is that of finding young transiting exoplanets.
Young transiting planets are hard to find because young stars are
rare, and mostly reside in crowded galactic plane \citep[see {\it e.g.},][]{Kharchenko_et_al_2013,piskunov_global_2018}.  If the
dispersed halos of nearby star clusters could be reliably identified,
this could expand the census of nearby young stars by up to a factor
of $10$, based on the expected fraction of stars thought to be lost
during gas dispersal.

The detection of dispersed stellar associations, although
possible before Gaia, has now reached a breakthrough pace \citep[{\it
e.g.},][]{bergond_gravitational_2001,zuckerman_young_2004,oh_comoving_2017,cantatgaudin_gaia_2018,gagne_banyan_XII_2018,gagne_banyan_XIII_2018,kounkel_apogee2_2018,zari_3d_2018,kounkel_untangling_2019,furnkranz_comabernbhr_2019}.
The resuls can be summarized as follows. On one end are coeval populations with no
discernable cores ({\it e.g.}, Psc Eri and $\mu$ Tau,
\citealt{meingast_psceri_2019,curtis_tess_2019,gagne_mutau_2020}).  On
the other are hierarchically structured populations with many over and
under-densities  \citep[{\it e.g.} the Sco-Cen and Vela OB2
associations][]{pecaut_star_2016,cantatgaudin_velaOB2_2019}.  Here,
we focus on the middle: low-density halos associated with a single
overdensity, typically an open cluster that was known before Gaia
\citep[see][]{kounkel_untangling_2019,kounkel_untanglingII_2020,meingast_2021}.
In some cases, these halos could correspond to tidal tails, as have been
reported in the Hyades \citep{meingast_hyades_2019,roser_hyades_2019},
the Ursa Major moving group \citep{gagne_lowmassUMA_2020}, and in Coma
Berenices \citep{tang_comaber_2019}.

One point of difficulty however is that
different clustering methods yield differing levels of both
sensitivity and precision \citep{hunt_clustering_2020}.  Using say a
Gaussian Mixture Model tautologically yields open clusters that
are elliptical \citep[{\it e.g.},][]{wallace_m4_2018}.  Some
unsupervised methods yield dispersed and asymmetric structures with
number densities down to 100 times less than the field around the same
regions (see {\it e.g.} \citealt{kounkel_untangling_2019}, hereafter
\citetalias{kounkel_untangling_2019}], and
\citealt{meingast_2021}, hereafter \citetalias{meingast_2021}).
Discrepancies between different clustering methods leads to
the question of how contaminated the resulting samples are,
and in some cases whether the reported structures are truly
coeval stellar populations.

We have recently been making TESS light curves of stars reported
to be members of such populations
(CDIPS, \citealt{bouma_cdipsI_2019}).  Our analysis of TESS Cycle 1
(Sectors 1-13) yielded light curves of \numsouthernuniqlcs candidate
cluster members in the Southern Ecliptic hemisphere, available on
MAST\footnote{\url{https://archive.stsci.edu/hlsp/cdips}}.

%; B3V at the turnoff
As part of this broader project, we focus here on a modest question:
in just a single rich southern open cluster, is the cluster halo truly
coeval with the core?  We chose NGC\,2516 for this analysis since it
is young ($\approx$120\,Myr) and sufficiently nearby
($d\approx400$\,pc) to facilitate photometric rotation measurements
using TESS, and some degree of spectroscopic analysis.  Its halo is
decidedly non-spherical: \citetalias{kounkel_untangling_2019} reported
its dimensions as $\approx20\times10\times350$\,pc.  We want
to know: is the halo real? To what extent can we use Gaia alone to
reliably identify age-dated needles in the haystack of field
stars?  What are the implications, observationally and theoretically,
if we can identify the dispersed halos of open clusters?

A brief
note on terminology.  Low-density stellar associations connected to a
dense population (the ``core'') have been described as ``halos'',
``strings'', ``coronae'', ``snakes'', and ``tidal tails'' \citep[{\it
	e.g.},][]{davenport_death_2010,kounkel_untangling_2019,roser_hyades_2019,tian_discovery_2020,meingast_2021}.
The latter term implies a particular model for the
formation of the association.  We opt to use ``halo'' because it is
concise and model agnostic.  It suffers however from the suggestion of
spherical symmetry.  The halo of NGC\,2516 is anything but.

Section~\ref{sec:gaia} summarizes the astrometric analyses of the Gaia
data that led to our interest in NGC\,2516.
Section~\ref{sec:agedate} age-dates the cluster's halo and core using
Gaia photometry (Section~\ref{subsec:hr}), TESS gyrochronology
(Section~\ref{subsec:tess}), and lithium depletion
(Section~\ref{subsec:lithium}).  In Section~\ref{sec:discussion} we
discuss the implications of our analysis for NGC\,2516 specifically
and stellar spin-down and open cluster evolution generally.
Section~\ref{sec:conclusion} gives our conclusions.

@@

Gaia-ESO selected {\it cluster candidates} to be observed with the
GIRAFFE and UVES spectrographs based on previously reported literature
members and publicly available optical and NIR photometry.  This
target selection approach included many stars in the spatial proximity
of previously known members.  Since the possible existence of the \cn\ 
halo was not known at the time of targetting, very few ``halo'' stars
are in their ultimate sample.  For \cn, \citet{randich_gaiaeso_2018}
ultimately reported stellar parameters (including lithium equivalent
widths and metallicities) for 796 candidate members.  Cross-matching
against our kinematic list of \nkinematic candidate members by position and
imposing a 0$\farcs$5 maximum separation limit yields 492 kinematic
members for which a Gaia-ESO spectrum was acquired and stellar
parameters were reported.  15 of these (all with $(Bp-Rp)_0 > 2.0$)
are spurious matches based on the Gaia color and GES effective
temperature, and we remove them.  This yields 477 stars, of which 436
are in the core, and 41 are in the halo.  This ratio is primarily
caused by the Gaia-ESO targetting selection function.

@@

% TODO: fix the fact that the figure currently only shows Randich+18
Figure~\ref{fig:lithium} shows the results.  At fixed stellar mass
(and age), the rapid rotators tend to show elevated lithium equivalent
widths. This effect is mostly apparent in the K dwarfs.  Similar
trends in the Pleiades were noted nearly three decades ago by
\citet{soderblom_evolution_1993}.  More recent studies have found
comparable results in the Pleiades, the Psc-Eri stream, and M\,35
\citep{bouvier_pleiades_lirot_2018,arancibia_2020,jeffries_m35_li_2020}.
The fact that we see the same trend across the core and halo of \cn\ 
{\it a)} supports the conclusion that the halo is coeval with the
core, and {\it b)} suggests that the lithium-rotation correlation is
not explained by environmental effects such as the density of the cluster
during its embedded phase, but is instead tied to physical processes
acting on all low-mass pre-main sequence stars.

### 2021-02-22 planet_paper/ms.tex
update figures
update paper
the introduction is no longer garbage

@@

\subsection{The future}:
- Detailed multi-abundance modelling.
  i.e,. Get spectra of other cluster stars using PFS, and compare the metallicities.
- 2$\farc$5 separate, multiband imager. (With LCOGT, 2m in Australia,
  z-band)->Do PSF fitting.
- K-band image. MagAO? Multicolor imager.
  E.g., Keck-NIRC2 (if it were in the North).
  ???What is the prediction for the G-K color???
- Maybe GROND -- has K-band. At La Silla. On 2.2m.  griz+K.

>>>>

\paragraph{Isn't the rotation period a little slow?}
Yes. But hot Jupiters affect the star's rotation, and it's hard to be
sure how much. Consider TOI 1431b, an Am-stype with vsini=8km/s
hosting a retrograde hot Jupiter. 
Generally speaking, retrograde systems will spin down the star (e.g.,
Anderson+2021), even if the planet is quickly ``realigned''.
The latter realignment would be needed in our case to understand TOI
1937Ab, since the observed orbit is prograde.

@@

\subsection{The future}:
- Detailed multi-abundance modelling.
  i.e,. Get spectra of other cluster stars using PFS, and compare the metallicities.
- 2$\farc$5 separate, multiband imager. (With LCOGT, 2m in Australia,
  z-band)->Do PSF fitting.

- K-band image. MagAO? Multicolor imager.
  E.g., Keck-NIRC2 (if it were in the North).
  ???What is the prediction for the G-K color???
  Doesn't seem all that promising. G-mag would predict something like
  M2.5V or M3.5V (0.40Msun or 0.30Msun from mamajek).
  At 100 Myr, a 0.20Msun star is 18Gcm vs 16Gcm (12\% inflated;
  Burrows+01). This is not a whole lot. And it is worse for more massive
  stars.
- Maybe GROND -- has K-band. At La Silla. On 2.2m.  griz+K.
  Going whole hog, maybe SPHERE could do it.
